openarchaeoCollaboration

### 2023-01-30 analysis/paper.Rmd
Minor formatting changes

## Introduction
In applying either physical or digital methods, archaeology has a long tradition of adopting common or 'off the shelf' tools to serve our own unique purposes, which are not typically unanticipated by their original designers. In other words, the apparatus of archaeology is an open system that relies on products and skillsets from outside the discipline, but which exhibit enough flexibility to be used in novel ways. For example, archaeologists are content to outsource the metallurgical skills needed to forge their trowels, and delegate the manufacture and acquisition of key electronic tools, like laptop computers and cameras, to the global supply chains and labour networks that produce consumer-grade products for purchase. However, an increasingly large number of archaeologists are taking it upon themselves to program their own software, rather than rely on generic and commercially-produced information processing systems. In fact, archaeologists are prolific computer programmers and have been coding a variety of functions and libraries that serve uniquely archaeological problems and purposes.

@@

(1) we identify _what_ kinds of software archaeologists are making;
(2) and evaluate _how_ archaeologists create these tools, with particular emphasis on practices of collaboration.

@@

## Background
- The notion of the archaeological toolbox, and its inclusion of digital tools and skillsets
  - The use of off-the-shelf tools like MS Access, Excel, Agisoft PhotoScan, Dropbox, etc
  - Link to related factors:
    - Lack of experienced programmers among archaeologists
    - The built-in flexibility of these tools
    - Ease of administrative purchasing through university departments
  - What changed?
    - Software is expensive, shift to subscription models, locking into a dependency on a commercial service provider
    - Archaeologists are better at computer programming, digital literacy on the rise, digital archaeologists grows into its own legitimate sub-discipline
    - Open source is cool, everyone has a github profile (even if not really)
      - (JR:) People are presumably guided towards GitHub (or at least git) via other routes, because it's seen as the standard platform for 'open' digital work. For example, the literature on reproducible compendiums recommends GitHub, and almost all R development work is now done there. And once there, you're sort of forced into engaging in a (particular set of) open source development. I know there's controversy about how dominant this has become and people who say it's not 'true' (free and) open source, and perhaps we could add to that the question of whether it's the right model for archaeology.

- About open source practices and their influence on archaeology
  - What is open source software and open source software development?
    - Free cost as an end-user
    - Free to manipulate and extend
    - Community-based support and development
  - Software as product vs software as collective process
    - Basically what distinguishes abandonware from thriving software ecosystems
    - Technical, legal and social notions of open
      - Technical: source code is open, use of a platform and systems that support software integration and implementations
      - Legal: open licenses, backing by the courts
      - Social: inviting community, meaningful contributions provided by users, common vision and values directing the project into future directions
    - Consider how these aspects of open have been implemented or prioritized by digital archaeologists in different ways
      - Technical: reproducible research, archiving and publishing
      - Legal: archiving and publishing, budgeting for projects
      - Social: ensuring that software is actually useful for general use and not just one-off implementation, coordinating to reduce redundant effort and not reinvent the wheel, writing documentation, making it fit within existing systems or infrastructures (i.e. allocation of credit, citing software directly versus articles that present software, sustainable development after funding runs out, etc)
    - This paper largely focuses on the social aspect of open, which is arguably the most important for ensuring a positive future for archaeological software development

@@

Please get in touch if you’re interested in adding to the list or doing a deep dive relating to your area of expertise!

### Tags and categories
Each record is also annotated with tags that describe the aspects of archaeological work that each tool contributes to.

@@

|Category|Kind|Scope|
|:-------|:---|:----|
|Packages and libraries|Software|Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.|
|Standalone software|Software|Software that may be operated without needing to first access an underlying platform.|
|Scripts|Software|Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.|
|Specifications, protocols and schemas|Documents|A formal data structure or framework intended to be used as a model.|
|Products|Documents|Stable outcomes of creative work.|
|Guides|Documents|An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.|
|Lists and datasets|Documents|A series of consistently organized observations assembled with purpose.|

@@

## Findings
### What archaeologists are making
Our first set of analyses were focused on these questions:

@@

---

Most projects included in open-archaeo are designed to be used as a package or library atop am existing ‘platform’ – for example a package for a programming language like R, or a plugin for an application like QGIS.

@@

---

@@

In the figure below, the darker grey lines are used to break up the bars by primary author, showing that many of the smaller platforms—for example Excel templates—are targeted by only one or two developers. The larger platforms tend to be more diverse.

@@

### Development processes
While somewhat informative, breaking down the distribution of categories and tags does not really tell us that much about _how_ archaeologists develop software.

@@

To address these questions, we augmented the core open-archaeo dataset with data obtained from GitHub’s public API.

@@

This does reduce our sample size slightly; about an eighth of projects don’t use version control, at least not publicly, but as yu can see in the figure below, amongst those that do, GitHub is by far the most popular host.

@@

As we can see, archaeologists have been using git from even before GitHub was launched in 2008. But they really began to get into it around 2014–2015, when we see an uptick in the rate of growth that has continued ever since.

Around this time we also see that GitHub starts being used to host documents and scripts. This may represent a recognition of GitHub’s ability to track things other than code, and a willingness to experiment with version control systems as a medium for disseminating work, especially work or manifestations of undervalued skills that may not necessarily be easily represented using traditional publication venues, in an open and somewhat nerdy way.

@@

This visualizes the ‘pulse’ of active projects.

@@

---

Perhaps unfortunately then, we also find that nearly 60% of projects have only a single contributor, and projects with more than a dozen are very rare indeed [<3%].

A “contribution”, in this case, is a commit, issue, comment applied by a user to a repository.

@@

### Formation of collaborative development communities
If we construct a network linking repositories by their common contributors, can we identify emerging networks of collaboration, or are individual projects still largely isolated? If there are networks, what characterises their links?

@@

With this, we can construct our three graphs: repo–user, repo–repo (by common users), and user–user (by common repos).

#### Repo–user

@@

#### Repo–repo

@@

#### User–user

@@

The vast majority of repositories have 1–3 contributors, with only a few distinguished by an active and diverse developer base.

Generally speaking, we believe that this is because people do not want to step on other people’s shoes by raising issues or intruding on other people’s projects.

This may relate to our positions as academics who hold different values relative to the designers of open source development environments, regarding how collaboration should occur, for example, when dealing with how projects and ideas are ‘owned’ by individuals or communities, and how work should be iteratively improved upon.

@@

Since most repositories use more than one language—if only one for the actual software, and another (e.g. HTML) for documentation—this data is presented as a table of languages and the number of bytes of code in that language detected in the repository.

@@

Looking at number of bytes puts C++ and the web languages (HTML, PHP, and Javascript) on top, but this is probably affected by the relative verboseness of these languages and the popularity of web-based documentation across implementation languages. The number of repositories is probably more reliable and shows the popularity of scientific programming languages—R and Python—followed by web languages and then compiled C and C++.

>>>>

## Introduction

In applying either physical or digital methods, archaeology has a long tradition of adopting common or 'off the shelf' tools to serve our own unique purposes, which are not typically unanticipated by their original designers. In other words, the apparatus of archaeology is an open system that relies on products and skillsets from outside the discipline, but which exhibit enough flexibility to be used in novel ways. For example, archaeologists are content to outsource the metallurgical skills needed to forge their trowels, and delegate the manufacture and acquisition of key electronic tools, like laptop computers and cameras, to the global supply chains and labour networks that produce consumer-grade products for purchase. However, an increasingly large number of archaeologists are taking it upon themselves to program their own software, rather than rely on generic and commercially-produced information processing systems. In fact, archaeologists are prolific computer programmers and have been coding a variety of functions and libraries that serve uniquely archaeological problems and purposes.

@@

(1) we identify *what* kinds of software archaeologists are making;
(2) and evaluate *how* archaeologists create these tools, with particular emphasis on practices of collaboration.

@@

## Background

-   The notion of the archaeological toolbox, and its inclusion of digital tools and skillsets
    -   The use of off-the-shelf tools like MS Access, Excel, Agisoft PhotoScan, Dropbox, etc
    -   Link to related factors:
        -   Lack of experienced programmers among archaeologists
        -   The built-in flexibility of these tools
        -   Ease of administrative purchasing through university departments
    -   What changed?
        -   Software is expensive, shift to subscription models, locking into a dependency on a commercial service provider
        -   Archaeologists are better at computer programming, digital literacy on the rise, digital archaeologists grows into its own legitimate sub-discipline
        -   Open source is cool, everyone has a github profile (even if not really)
            -   (JR:) People are presumably guided towards GitHub (or at least git) via other routes, because it's seen as the standard platform for 'open' digital work. For example, the literature on reproducible compendiums recommends GitHub, and almost all R development work is now done there. And once there, you're sort of forced into engaging in a (particular set of) open source development. I know there's controversy about how dominant this has become and people who say it's not 'true' (free and) open source, and perhaps we could add to that the question of whether it's the right model for archaeology.
-   About open source practices and their influence on archaeology
    -   What is open source software and open source software development?
        -   Free cost as an end-user
        -   Free to manipulate and extend
        -   Community-based support and development
    -   Software as product vs software as collective process
        -   Basically what distinguishes abandonware from thriving software ecosystems
        -   Technical, legal and social notions of open
            -   Technical: source code is open, use of a platform and systems that support software integration and implementations
            -   Legal: open licenses, backing by the courts
            -   Social: inviting community, meaningful contributions provided by users, common vision and values directing the project into future directions
        -   Consider how these aspects of open have been implemented or prioritized by digital archaeologists in different ways
            -   Technical: reproducible research, archiving and publishing
            -   Legal: archiving and publishing, budgeting for projects
            -   Social: ensuring that software is actually useful for general use and not just one-off implementation, coordinating to reduce redundant effort and not reinvent the wheel, writing documentation, making it fit within existing systems or infrastructures (i.e. allocation of credit, citing software directly versus articles that present software, sustainable development after funding runs out, etc)
        -   This paper largely focuses on the social aspect of open, which is arguably the most important for ensuring a positive future for archaeological software development

@@

Please get in touch if you're interested in adding to the list or doing a deep dive relating to your area of expertise!

### Tags and categories

Each record is also annotated with tags that describe the aspects of archaeological work that each tool contributes to.

@@

| Category                              | Kind      | Scope                                                                                                                                                                                                        |
|:--------------------------------------|:----------|:-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| Packages and libraries                | Software  | Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.                                                                                   |
| Standalone software                   | Software  | Software that may be operated without needing to first access an underlying platform.                                                                                                                        |
| Scripts                               | Software  | Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation. |
| Specifications, protocols and schemas | Documents | A formal data structure or framework intended to be used as a model.                                                                                                                                         |
| Products                              | Documents | Stable outcomes of creative work.                                                                                                                                                                            |
| Guides                                | Documents | An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.                                                                                          |
| Lists and datasets                    | Documents | A series of consistently organized observations assembled with purpose.                                                                                                                                      |

@@

## Findings

### What archaeologists are making

Our first set of analyses were focused on these questions:

@@

------------------------------------------------------------------------

Most projects included in open-archaeo are designed to be used as a package or library atop am existing 'platform' -- for example a package for a programming language like R, or a plugin for an application like QGIS.

@@

------------------------------------------------------------------------

@@

In the figure below, the darker grey lines are used to break up the bars by primary author, showing that many of the smaller platforms---for example Excel templates---are targeted by only one or two developers. The larger platforms tend to be more diverse.

@@

### Development processes

While somewhat informative, breaking down the distribution of categories and tags does not really tell us that much about *how* archaeologists develop software.

@@

To address these questions, we augmented the core open-archaeo dataset with data obtained from GitHub's public API.

@@

This does reduce our sample size slightly; about an eighth of projects don't use version control, at least not publicly, but as yu can see in the figure below, amongst those that do, GitHub is by far the most popular host.

@@

As we can see, archaeologists have been using git from even before GitHub was launched in 2008. But they really began to get into it around 2014--2015, when we see an uptick in the rate of growth that has continued ever since.

Around this time we also see that GitHub starts being used to host documents and scripts. This may represent a recognition of GitHub's ability to track things other than code, and a willingness to experiment with version control systems as a medium for disseminating work, especially work or manifestations of undervalued skills that may not necessarily be easily represented using traditional publication venues, in an open and somewhat nerdy way.

@@

This visualizes the 'pulse' of active projects.

@@

------------------------------------------------------------------------

Perhaps unfortunately then, we also find that nearly 60% of projects have only a single contributor, and projects with more than a dozen are very rare indeed [\<3%].

A "contribution", in this case, is a commit, issue, comment applied by a user to a repository.

@@

### Formation of collaborative development communities

If we construct a network linking repositories by their common contributors, can we identify emerging networks of collaboration, or are individual projects still largely isolated? If there are networks, what characterises their links?

@@

With this, we can construct our three graphs: repo--user, repo--repo (by common users), and user--user (by common repos).

#### Repo--user

@@

#### Repo--repo

@@

#### User--user

@@

The vast majority of repositories have 1--3 contributors, with only a few distinguished by an active and diverse developer base.

Generally speaking, we believe that this is because people do not want to step on other people's shoes by raising issues or intruding on other people's projects.

This may relate to our positions as academics who hold different values relative to the designers of open source development environments, regarding how collaboration should occur, for example, when dealing with how projects and ideas are 'owned' by individuals or communities, and how work should be iteratively improved upon.

@@

Since most repositories use more than one language---if only one for the actual software, and another (e.g. HTML) for documentation---this data is presented as a table of languages and the number of bytes of code in that language detected in the repository.

@@

Looking at number of bytes puts C++ and the web languages (HTML, PHP, and Javascript) on top, but this is probably affected by the relative verboseness of these languages and the popularity of web-based documentation across implementation languages. The number of repositories is probably more reliable and shows the popularity of scientific programming languages---R and Python---followed by web languages and then compiled C and C++.
